When Risk Hits the Road: Does Danger Drive a Return?

### 2024-12-02 paper/paper.qmd
finished the first draft of result section of the paper

Decisions surrounding motorbike ownership and usage carry significant implications for personal safety and financial liability. Recent statistics highlight the increased risks associated with owning and riding motorbikes, including a heightened likelihood of theft and collisions compared to other vehicles [@Yasmin2016]. These risks are influenced by various factors such as geographic location, road conditions, time of day, and type of motorbike, making it essential to develop tools that effectively assess and mitigate these dangers.[can add one more sentence + citations]

@@

This analysis confirms and extends three key findings: (1) material properties significantly influence motor efficiency, with high-conductivity materials yielding higher energy savings; (2) geometric optimization of motor components enhances torque generation and reduces energy losses; and (3) operational conditions, such as load and temperature, exhibit nonlinear effects, underscoring the need for adaptive motor control strategies. Furthermore, our projections suggest that integrating advanced materials and control systems could improve motor efficiency by up to 15% by 2050, aligning with global energy transition goals [To be updated..].

@@

[Libraries To be updated...]

@@

[Add variable types and trends]

@@

The outcome variable of this analysis is the Risk Index, a composite metric that integrates the probabilities of two underlying events: theft and collision.

Need two graphs
```{r}

@@

@@

@@

@@

#### Hour of the Day
#### Day of the Week
#### Premises Type

@@

#### Neighborhood ID
#### Road Surface Condition
#### Lighting Condition
#### Traffic Control
#### Time of Day
#### Visibility

@@

### Collision Probability Model
A logistic regression model was used to predict the likelihood of severe collisions P(Collision) based on several predictors. The log-odds of the collision probability are modeled as:
\

\begin{align*}
\log\left(\frac{P(\text{Collision})}{1 - P(\text{Collision})}\right) &= \beta_0 + \beta_1 \cdot \text{HOOD}_{158} + \beta_2 \cdot \text{Road Surface Condition} + \beta_3 \cdot \text{Lighting Condition} \\
&\quad + \beta_4 \cdot \text{Traffic Control} + \beta_5 \cdot \text{Road Visibility} + \beta_6 \cdot \text{Time of Day} + \epsilon
\end{align*}

The model prediction utilizes the following predictor variables:

@@

\begin{align*}
\beta_i \text{ to each variable, enabling the calculation of collision probability }\\ 
P(\text{Collision})\text{ under specific environmental and situational conditions.} 
\end{align*}

### Risk Index Calculation

@@

\begin{align*}
\text{Risk Index} = w_1\cdot P(\text{Collision}) + w_2\cdot T
\end{align*}

Weights are defined as:
\begin{align*}
w_1 = 0.7, \quad w_2 = 0.3
\end{align*}
These reflect the relative importance of collision and theft risks, emphasizing collision severity due to its greater immediate impact.

@@

### Theft Sub-Index [add one sentence of validation]

@@

### Composite Risk Index [Add one sentence of assumption]

@@

In the collision model:
- The model estimates the probability of a severe collision (\(P(\text{Collision})\)) given environmental and situational predictors. The response variable (`severity`) is defined as:
  \begin{align*}
  P(\text{Collision}) = \frac{\exp(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k)}{1 + \exp(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k)}
  \end{align*}
  where:
  \begin{align*}
  \beta_0 &\text{ is the intercept,} \\
  \beta_k &\text{ are the coefficients, and} \\
  X_k &\text{ are the predictor variables.}
  \end{align*}

@@

\begin{align*}
\text{logit} = \log\left(\frac{P}{1 - P}\right)
\end{align*}

where P represents the predicted probability of collision from the logistic regression model plotted against the continuous predictors. These plots are intended to visualize the relationship between each predictor and the logit of the outcome variable, providing insight into whether the relationship is approximately linear.

```{r}
#| label: fig-linear-time-of-day
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Logit Plot for Time of Day in the Collision Probability Model"
#| fig-pos: H
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(here)

@@

# Smoothed scatter plot for time_of_day vs logit
ggplot(collision_data, aes(x = time_of_day, y = logit)) +
  geom_point(alpha = 0.5, color = "darkgrey") +
  geom_smooth(method = "loess", color = "blue", fill = "lightblue") +
  coord_cartesian(ylim = c(-5, 5)) +  # Restrict the y-axis range
  labs(
    x = "Time of Day",
    y = "Logit"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")  # Center title and adjust size
  )

@@

Despite this overlap, both predictors are retained in the model because they provide distinct and meaningful contributions to understanding collision risk. Visibility Conditions directly reflects environmental factors like fog, heavy rain, or low light, which impair drivers' ability to see hazards. Conversely, Road Conditions` account for the physical state of the driving surface, such as wet, icy, or damaged roads, which influence vehicle handling and stopping distance. Together, these variables capture complementary aspects of collision risk, ensuring that the model provides a comprehensive assessment.

@@

Time-of-Day and Day-of-Week Analysis of the Risk Index

Plot Ideas:
Bar: Show Risk Index variations across hours of the day and days of the week, highlighting theft and collision components separately.

Time Series Plot: Display Risk Index trends over a longer period, revealing potential seasonal effects.

```{r}
```

@@

@@

Exploring How Environmental and Situational Factors Impact the Risk Index

Plot Ideas:
Violin Plot: Show the Risk Index distribution under varying road surface conditions (e.g., dry, wet, icy).

Faceted Plot: Break down the Risk Index by lighting conditions (e.g., daylight, artificial light) or traffic control types.

>>>>

Decisions surrounding motorbike ownership and usage carry significant implications for personal safety and financial liability. Recent statistics highlight the increased risks associated with owning and riding motorbikes, including a heightened likelihood of theft and collisions compared to other vehicles [@Yasmin2016]. These risks are influenced by various factors such as geographic location, road conditions, time of day, and type of motorbike, making it essential to develop tools that effectively assess and mitigate these dangers.\[can add one more sentence + citations\]

@@

This analysis confirms and extends three key findings: (1) material properties significantly influence motor efficiency, with high-conductivity materials yielding higher energy savings; (2) geometric optimization of motor components enhances torque generation and reduces energy losses; and (3) operational conditions, such as load and temperature, exhibit nonlinear effects, underscoring the need for adaptive motor control strategies. Furthermore, our projections suggest that integrating advanced materials and control systems could improve motor efficiency by up to 15% by 2050, aligning with global energy transition goals \[To be updated..\].

@@

\[Libraries To be updated...\]

@@

\[Add variable types and trends\]

@@

The outcome variable of this analysis is the Risk Index, a composite metric that integrates the probabilities of two underlying events: theft and collision.The Risk Index integrates two underlying components: the theft component, derived from proportional sub-indexes for temporal and spatial factors, and the collision probability, estimated using a logistic regression model. By combining these elements, the Risk Index provides a unified measure of risk, enabling the identification of high-risk scenarios and areas. This section examines the distribution of the Risk Index through visualizations to reveal key patterns, variations, and potential drivers of motorbike-related risks. Two graphs are presented: (1) a histogram with density overlay to illustrate the overall distribution of the Risk Index, and (2) a boxplot by neighborhood to highlight spatial differences in risk.

```{r}
#| label: fig-risk-index-histogram
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Distribution of the Overall Risk Index"
#| fig-pos: H
#| fig-width: 8 
#| fig-height: 4

# Load necessary libraries
library(tidyverse)
library(here)

# Load the final Risk Index dataset
final_risk_index <- read_csv(here("data", "02-analysis_data", "final_risk_index.csv"))

# Plot a histogram with density overlay
ggplot(final_risk_index, aes(x = risk_index)) +
  geom_histogram(aes(y = ..density..), bins = 40, fill = "darkgrey", alpha = 0.6, color = "black") +
  geom_density(color = "darkred", size = 1) +
  labs(

    x = "Risk Index",
    y = "Density"
  ) +
  theme_minimal()

@@

@fig-risk-index-histogram shows the distribution of the Risk Index across all observations in the dataset. The Risk Index exhibits a unimodal distribution, skewed slightly to the left, with the majority of values concentrated between 0.45 and 0.65. This indicates that most motorbike-related risks fall within a moderate range. The density curve overlay reveals a smooth progression in risk levels, with the peak occurring around a Risk Index value of 0.55, suggesting that this is the most common level of composite risk. The left tail, representing lower risk levels, is relatively small, while the right tail, corresponding to higher risks, extends further, indicating the presence of a smaller number of high-risk cases.

The skewness and spread of the distribution highlight the variability in the combined risks of theft and collision. The extended tail on the higher end of the Risk Index suggests that certain environmental or situational factors disproportionately elevate risks in specific cases. This insight can inform targeted interventions, focusing on the outliers with elevated Risk Index values to mitigate the most critical risks.

Building on the distribution of the Risk Index shown in the histogram, this box plot further dissects the data by grouping neighborhoods into risk categories based on their average Risk Index values. Neighborhoods are classified into three categories—Low Risk, Moderate Risk, and High Risk—using quartiles.

```{r}
#| label: fig-risk-index-boxplot
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Risk Index Distribution by Risk Category"
#| fig-pos: H

library(ggplot2)
library(dplyr)

risk_index_data <- read_csv(here("data", "02-analysis_data", "final_risk_index.csv"))

# Assume `risk_index_data` contains the `risk_index` values for each neighborhood
# Group neighborhoods by average Risk Index
grouped_data <- risk_index_data %>%
  group_by(hood_158) %>%
  summarize(avg_risk_index = mean(risk_index, na.rm = TRUE)) %>%
  mutate(risk_category = case_when(
    avg_risk_index <= quantile(avg_risk_index, 0.25) ~ "Low Risk",
    avg_risk_index > quantile(avg_risk_index, 0.25) & avg_risk_index <= quantile(avg_risk_index, 0.75) ~ "Moderate Risk",
    avg_risk_index > quantile(avg_risk_index, 0.75) ~ "High Risk"
  ))

# Add risk categories directly to the `risk_index_data`
risk_index_data <- risk_index_data %>%
  left_join(grouped_data %>% select(hood_158, risk_category), by = "hood_158")

# Box plot of Risk Index by Risk Category with custom colors
ggplot(risk_index_data, aes(x = risk_category, y = risk_index, fill = risk_category)) +
  geom_boxplot(outlier.color = "darkred", outlier.shape = 16, alpha = 0.7) +
  scale_fill_manual(
    values = c("Low Risk" = "#603601", "Moderate Risk" = "#3a6a91", "High Risk" = "#cfb886")
  ) +
  labs(
    x = "Risk Category",
    y = "Risk Index"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

@@

@fig-risk-index-boxplot highlights the distribution of the Risk Index across three defined categories: High Risk, Low Risk and Moderate Risk. Each category represents a grouping of neighborhoods based on their average Risk Index values. The plot reveals distinct differences in the distribution of the Risk Index between these categories.

The High Risk category exhibits a higher median Risk Index, with a relatively narrow interquartile range (IQR), indicating consistent high-risk values across neighborhoods in this group. Conversely, the Low Risk category has a significantly lower median, with a similarly narrow IQR, suggesting stable low-risk conditions in these neighborhoods. The Moderate Risk category shows greater variability in its distribution, with a wider IQR and overlapping values with both High Risk and Low Risk categories. This suggests that neighborhoods in the Moderate Risk group exhibit diverse risk profiles, likely influenced by varying environmental and situational factors.

@@

The predictor variables in this study are organized into two distinct models: Theft Index and Collision Probability, each designed to capture and explain critical aspects of theft and collision risks, respectively. @fig-tree provide a visual overview of the hierarchical relationships between the predictor variables and the overall risk framework, offering a structured understanding of the factors contributing to these incidents

```{r}
#| label: fig-tree
#| fig-cap: "Hierarchical relationship between risk factors contributing to theft and collision probabilities."
#| echo: false
#| warning: false
#| message: false
#| fig-width: 8 # Increased width for better text display
#| fig-height: 4  # Adjusted height for spacing

# Load required libraries
library(tidygraph)
library(ggraph)

# Define the tree structure as a dataframe
tree_data <- data.frame(
  from = c("Risk Index", "Risk Index", "Theft Index", "Theft Index", "Theft Index",
           "Collision Probability", "Collision Probability", "Collision Probability", 
           "Collision Probability", "Collision Probability", "Collision Probability"),
  to = c("Theft Index", "Collision Probability", "Hour of the Day", "Day of the Week", 
         "Premises Type", "Neighborhood ID", "Road Surface Condition", 
         "Lighting Condition", "Traffic Control", "Time of Day", "Road Visibility")
)

# Convert to graph object
tree_graph <- as_tbl_graph(tree_data)

# Plot the tree diagram with horizontal layout and curved lines
ggraph(tree_graph, layout = "dendrogram", circular = FALSE) +
  geom_edge_bend(color = "gray", curvature = 0.5, arrow = arrow(length = unit(4, "mm")), end_cap = circle(4, "mm")) +
  geom_node_point(size = 6, color = "grey") +
  geom_node_text(
    aes(label = name, hjust = ifelse(name == "Risk Index", 1.3, -0.2)), # Conditional hjust
    vjust = -0.5, size = 4, color = "black", angle = 0
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  ) +
  coord_flip()

@@

The Theft Index focuses on the temporal and locational characteristics that influence theft occurrences. By incorporating variables such as Hour of the Day, Day of the Week, and Premises Type, this model identifies patterns tied to specific times and locations, highlighting when and where thefts are most likely to occur.

#### Hour of the Day

#### Day of the Week

#### Premises Type

@@

The Collision Probability model, on the other hand, examines situational and environmental conditions affecting the likelihood of collisions. Variables such as Neighborhood ID, Road Surface Condition, Lighting Condition, Traffic Control, Time of Day, and Visibility offer insights into the contextual factors that contribute to traffic incidents, capturing the dynamic interplay between environmental conditions and human interactions.

#### Neighborhood ID

#### Road Surface Condition

#### Lighting Condition

#### Traffic Control

#### Time of Day

#### Visibility

@@

### Collision Probability Model

A logistic regression model was used to predict the likelihood of severe collisions P(Collision) based on several predictors. The log-odds of the collision probability are modeled as:

```{=tex}
\begin{align*}
\log\left(\frac{P(\text{Collision})}{1 - P(\text{Collision})}\right) &= \beta_0 + \beta_1 \cdot \text{HOOD}_{158} + \beta_2 \cdot \text{Road Surface Condition} + \beta_3 \cdot \text{Lighting Condition} \\
&\quad + \beta_4 \cdot \text{Traffic Control} + \beta_5 \cdot \text{Road Visibility} + \beta_6 \cdot \text{Time of Day} + \epsilon
\end{align*}
```
The model prediction utilizes the following predictor variables:

@@

```{=tex}
\begin{align*}
\beta_i \text{ to each variable, enabling the calculation of collision probability }\\ 
P(\text{Collision})\text{ under specific environmental and situational conditions.} 
\end{align*}
```
### Risk Index Calculation

@@

```{=tex}
\begin{align*}
\text{Risk Index} = w_1\cdot P(\text{Collision}) + w_2\cdot T
\end{align*}
```
Weights are defined as: \begin{align*}
w_1 = 0.7, \quad w_2 = 0.3
\end{align*} These reflect the relative importance of collision and theft risks, emphasizing collision severity due to its greater immediate impact.

@@

### Theft Sub-Index \[add one sentence of validation\]

@@

### Composite Risk Index \[Add one sentence of assumption\]

@@

In the collision model: - The model estimates the probability of a severe collision ((P(\text{Collision}))) given environmental and situational predictors. The response variable (`severity`) is defined as: \begin{align*}
  P(\text{Collision}) = \frac{\exp(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k)}{1 + \exp(\beta_0 + \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_k X_k)}
  \end{align*} where: \begin{align*}
  \beta_0 &\text{ is the intercept,} \\
  \beta_k &\text{ are the coefficients, and} \\
  X_k &\text{ are the predictor variables.}
  \end{align*}

@@

```{=tex}
\begin{align*}
\text{logit} = \log\left(\frac{P}{1 - P}\right)
\end{align*}
```
where P represents the predicted probability of collision from the logistic regression model plotted against the continuous predictors. These plots are intended to visualize the relationship between each predictor and the logit of the outcome variable, providing insight into whether the relationship is approximately linear.

```{r}
#| label: fig-linear-time-of-day
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Logit Plot for Time of Day in the Collision Probability Model"
#| fig-pos: H
#| fig-width: 8
#| fig-height: 3.5

# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(here)

@@

# Smoothed scatter plot for time_of_day vs logit
ggplot(collision_data, aes(x = time_of_day, y = logit)) +
  geom_point(alpha = 0.5, color = "darkgrey") +
  geom_smooth(method = "loess", color = "darkred", fill = "lightblue") +
  coord_cartesian(ylim = c(-5, 5)) +  # Restrict the y-axis range
  labs(
    x = "Time of Day",
    y = "Logit"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")  # Center title and adjust size
  )

@@

Despite this overlap, both predictors are retained in the model because they provide distinct and meaningful contributions to understanding collision risk. Visibility Conditions directly reflects environmental factors like fog, heavy rain, or low light, which impair drivers' ability to see hazards. Conversely, Road Conditions\` account for the physical state of the driving surface, such as wet, icy, or damaged roads, which influence vehicle handling and stopping distance. Together, these variables capture complementary aspects of collision risk, ensuring that the model provides a comprehensive assessment.

@@

Understanding the temporal dynamics of the Risk Index provides critical insights into when risks are most elevated, helping to inform motorbike owners and policymakers about high-risk periods. This section examines variations in the Risk Index over time of a day, and utilizes a time series analysis to reveal potential seasonal effects and long-term trends. These analyses provide a comprehensive view of how risks evolve temporally, highlighting critical periods for intervention.

### Time-of-Day Analysis

```{r}
#| label: fig-risk-index-timeofday
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Temporal Trends: Risk Index Variations by Hour of the Day"
#| fig-pos: H
#| fig-width: 8
#| fig-height: 4

# Load necessary data
final_risk_data <- read_csv(here("data", "02-analysis_data", "final_risk_index.csv"))

cleaned_traffic_data <- read_csv(here("data", "02-analysis_data", "cleaned_traffic_data.csv"))

collision_risk_scores <- read_csv(here("data", "02-analysis_data", "collision_risk_scores.csv"))

# Convert 'id' to character in both datasets if necessary
final_risk_data$id <- as.character(final_risk_data$id)
cleaned_traffic_data$id <- as.character(cleaned_traffic_data$id)
collision_risk_scores$id <- as.character(collision_risk_scores$id)

# Merge the final risk index with the traffic data based on 'id'
merged_data <- left_join(final_risk_data, cleaned_traffic_data, by = "id")

# Merge the collision risk scores into the merged data
merged_data <- left_join(merged_data, collision_risk_scores, by = "id")

# Create 'time_of_day_bin' based on 'time_of_day' in cleaned_traffic_data
merged_data$time_of_day_bin <- case_when(
  merged_data$time_of_day >= 0 & merged_data$time_of_day < 600 ~ "Early Morning",
  merged_data$time_of_day >= 600 & merged_data$time_of_day < 1200 ~ "Morning",
  merged_data$time_of_day >= 1200 & merged_data$time_of_day < 1800 ~ "Afternoon",
  merged_data$time_of_day >= 1800 & merged_data$time_of_day < 2400 ~ "Evening"
)

# Group by 'time_of_day_bin' and calculate average values for risk index and collision index
summary_data <- merged_data %>%
  group_by(time_of_day_bin) %>%
  summarize(
    avg_risk_index = mean(risk_index, na.rm = TRUE),
    avg_collision_index = mean(collision_probability
, na.rm = TRUE)
  )

# Theft hour indexes (24-hour data points)
theft_hour_indexes <- c(0.005335554, 0.003690682, 0.002343943, 0.002158895, 0.001572909,
                        0.002415906, 0.006353319, 0.012295419, 0.018587055, 0.02577309,
                        0.026174027, 0.025526359, 0.025248787, 0.024107657, 0.021969323,
                        0.020334731, 0.018196398, 0.016448721, 0.016479562, 0.015400115,
                        0.013806645, 0.01254215, 0.009211284, 0.007360803)

# Create a mapping of the theft hour indexes to the time of day bins
theft_hour_bins <- c("Early Morning", "Early Morning", "Early Morning", "Early Morning", "Early Morning", 
                     "Morning", "Morning", "Morning", "Morning", "Morning", 
                     "Morning", "Morning", "Morning", "Afternoon", "Afternoon", 
                     "Afternoon", "Afternoon", "Afternoon", "Afternoon", 
                     "Evening", "Evening", "Evening", "Evening", "Evening")

# Create a dataframe for the theft data, with time_of_day_bin mapped accordingly
theft_hour_data <- data.frame(
  time_of_day_bin = theft_hour_bins,
  theft_index = theft_hour_indexes
)

# Summarize theft data by time_of_day_bin to match the bin structure used in the plot
theft_summary_data <- theft_hour_data %>%
  group_by(time_of_day_bin) %>%
  summarize(avg_theft_index = mean(theft_index))

# Merge with existing summary data for risk and collision indexes
final_plot_data <- left_join(summary_data, theft_summary_data, by = "time_of_day_bin")

# Plot: Risk Index, Collision Index, and Theft Hour Index by Time of Day Bin
ggplot(final_plot_data, aes(x = time_of_day_bin)) +
  geom_bar(aes(y = avg_risk_index), stat = "identity", fill = "#967145") +
  geom_line(aes(y = avg_collision_index), group = 1, color = "darkred", size = 1) + # Collision index line
  geom_text(aes(y = avg_risk_index, label = round(avg_risk_index, 2)), vjust = -0.5) +  # Add labels for risk index
  geom_text(aes(y = avg_collision_index, label = round(avg_collision_index, 2)), vjust = 1.5, color = "black") +  # Add labels for collision index
  geom_line(aes(y = avg_theft_index * 10), color = "green", size = 1, linetype = "dashed") +  # Theft index line (scaled by 100)
  geom_point(aes(y = avg_theft_index * 10), color = "green", size = 2) +  # Theft index points
  labs(

    x = "Time of Day Bin",
    y = "Average Risk Index"
  ) +
  theme_minimal()

@@

@fig-risk-index-timeofday illustrates how the total Risk Index varies throughout the day. This Risk Index is composed of two key components: collision risk and theft risk. The bar heights represent the overall Risk Index, while the red line indicates collision risk and the green line reflects theft risk.To better visualize the theft risk, the values have been multiplied by 10, as they are typically quite small (less than 1/3 of the total Hour Risk Index throughout the day).

The collision risk (red line) shows significant peaks during commuting hours—between 7–9 AM and 5–7 PM—corresponding to high traffic volumes, which indicates a direct relationship between collision risk and traffic patterns. The theft risk (green line), which has been scaled for better visualization, remains consistently low throughout the day, showing no significant variation across time bins. This suggests that theft risk is relatively stable, regardless of the time of day. The total Risk Index, represented by the bar heights, shows slight fluctuations, with the highest values occurring during morning hours. This is likely due to the heightened collision risk in the morning, which has a greater influence on the overall risk during these periods.

### Time Series Analysis

This time series analysis explores how the Risk Index evolves over time, offering a long-term view across the years and a more focused examination of a specific year. By visualizing these temporal trends, we can reveal potential seasonal effects, fluctuations, and patterns that might not be immediately apparent in a snapshot of data. The following section presents the trends in the Risk Index over both the full timespan (2006-2021) and the specific year 2020.

#### Longterm

```{r}
#| label: fig-risk-index-timeseries-longterm
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Temporal Trends: Risk Index Variations Over Time with Rolling Average"
#| fig-pos: H
#| fig-width: 8
#| fig-height: 3.5

# Load necessary libraries
library(tidyverse)
library(lubridate)
library(here)

# Load the cleaned traffic data
cleaned_traffic_data <- read_csv(here("data", "02-analysis_data", "cleaned_traffic_data.csv"))

# Load the final risk index if not already in `cleaned_traffic_data`
if (!"risk_index" %in% colnames(cleaned_traffic_data)) {
  final_risk_data <- read_csv(here("data", "02-analysis_data", "final_risk_index.csv"))
  cleaned_traffic_data <- cleaned_traffic_data %>%
    left_join(final_risk_data, by = "id")
}

# Convert `date` to Date format if not already
cleaned_traffic_data <- cleaned_traffic_data %>%
  mutate(date = as.Date(report_date))

# Aggregate data by date to calculate the average Risk Index
risk_trends <- cleaned_traffic_data %>%
  group_by(date) %>%
  summarize(avg_risk_index = mean(risk_index, na.rm = TRUE))

# Create a 6-month rolling average for clearer trend
risk_trends <- risk_trends %>%
  arrange(date) %>%
  mutate(rolling_avg = zoo::rollmean(avg_risk_index, 6, fill = NA, align = "center"))

# Create the time series plot for the entire period
ggplot(risk_trends, aes(x = date, y = avg_risk_index)) +
  geom_line(color = "#967145", size = 1, alpha = 0.7) +
  geom_line(aes(x = date, y = rolling_avg), color = "grey", size = 1, linetype = "dashed") +
  labs(

    x = "Year",
    y = "Average Risk Index"
  ) +
  scale_x_date(
    date_breaks = "2 years", 
    date_labels = "%Y"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12)
  )

@fig-risk-index-timeseries-longterm provides a broad overview of the Risk Index over the entire period from 2006 to 2021, with the average Risk Index for each day represented by the blue line. To facilitate clearer interpretation, a 6-month rolling average is included, shown by the dashed red line. The plot highlights long-term trends in risk levels, showing periods of heightened and reduced risk. Although there are some fluctuations, the rolling average smooths these variations, enabling us to observe the general trend over time. The x-axis labels every three years for better clarity, making it easier to track shifts in risk levels. From this, we can discern if certain years or periods saw notable increases or decreases in the Risk Index, possibly suggesting underlying external factors like policy changes or events that affected the risk.

#### Year 2020

```{r}
#| label: fig-risk-index-timeseries-year
#| echo: false
#| warning: false
#| message: false
#| fig-cap: "Risk Index Variations in 2020"
#| fig-pos: H
#| fig-width: 8.5
#| fig-height: 3.5

# Filter data for a specific year (e.g., 2020)
risk_trends_2020 <- cleaned_traffic_data %>%
  filter(year(date) == 2020) %>%
  group_by(date) %>%
  summarize(avg_risk_index = mean(risk_index, na.rm = TRUE))

risk_trends_2020 <- risk_trends_2020 %>%
  mutate(month = month(date, label = TRUE))

ggplot(risk_trends_2020, aes(x = date, y = avg_risk_index)) +
  geom_line(color = "#66471d", size = 0.8) +
  facet_wrap(~month, scales = "free_x") +  # Facet by month
  labs(
    x = "Month",
    y = "Average Risk Index"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12)
  )

@fig-risk-index-timeseries-year provides a detailed, month-by-month breakdown of the Risk Index for the year 2020. This faceted layout offers a clear view of short-term fluctuations and potential seasonal patterns by presenting trends for each month separately. The graph highlights variations in risk levels across the months, making it easier to identify distinct periods of higher or lower risk. These trends may correspond to factors such as weather conditions, holidays, or other contextual events. For instance, spikes or dips in risk within specific months could reflect changes in traffic volume, seasonal behaviors, or other dynamic factors influencing motorbike-related incidents throughout the year.

@@

Understanding how environmental and situational factors contribute to variations in the Risk Index is crucial for identifying the conditions under which motorbike risks are heightened. This section explores how different factors, such as road surface conditions, lighting conditions, and traffic control types, influence the Risk Index. By breaking down the data into categories based on these factors, we can reveal specific conditions that might increase the likelihood of accidents or theft. The following visualizations provide insight into the distribution of risk under varying environmental conditions, highlighting the factors that contribute to motorbike-related incidents.

To better understand how environmental factors influence the Risk Index, we examine the distribution of the Risk Index under various road surface conditions. This analysis explores whether different surface types—such as dry, wet, icy, and others—affect the level of risk associated with traffic incidents. The following violin plot visually compares the Risk Index across different road surface conditions, providing insight into how these factors may contribute to the overall risk.

```{r}
#| label: fig-risk-index-lighting-conditions
#| fig-cap: "Risk Index by Lighting Conditions"
#| fig-subcap: ["Low Light (Dark & Artificial)", "Dawn and Artificial Light", "Bright Daylight and Artificial Light", "Other Lighting Conditions"]
#| echo: false
#| warning: false
#| message: false
#| layout-ncol: 2
#| layout-nrow: 2

# Load necessary libraries
library(ggplot2)
library(dplyr)

# Load the data
final_risk_data <- read_csv(here("data", "02-analysis_data", "final_risk_index.csv"))

cleaned_traffic_data <- read_csv(here("data", "02-analysis_data", "cleaned_traffic_data.csv"))

collision_risk_scores <- read_csv(here("data", "02-analysis_data", "collision_risk_scores.csv"))

# Convert 'id' to character in both datasets if necessary
final_risk_data$id <- as.character(final_risk_data$id)
cleaned_traffic_data$id <- as.character(cleaned_traffic_data$id)
collision_risk_scores$id <- as.character(collision_risk_scores$id)

# Merge the final risk index with the traffic data based on 'id'
merged_data <- left_join(final_risk_data, cleaned_traffic_data, by = "id")

# Merge the collision risk scores into the merged data
merged_data <- left_join(merged_data, collision_risk_scores, by = "id")

# Preprocess lighting conditions: Group 'NA' as 'Other' and create lighting condition categories for the full dataset
merged_data$lighting_conditions <- as.character(merged_data$lighting_conditions)  # Ensure it's a character vector

# Convert "NA" (actual value in the data) to "Other"
merged_data$lighting_conditions[merged_data$lighting_conditions == "NA"] <- "Other"

# Group lighting conditions into four categories, including "Other"
merged_data$lighting_condition_group <- case_when(
  merged_data$lighting_conditions %in% c("Dark", "Dark, Artificial") ~ "Low Light (Dark & Artificial)",
  merged_data$lighting_conditions %in% c("Dawn", "Dawn, Artificial") ~ "Dawn and Artificial Light",
  merged_data$lighting_conditions %in% c("Daylight", "Daylight, Artificial") ~ "Bright Daylight and Artificial Light",
  merged_data$lighting_conditions == "Other" ~ "Other Lighting Conditions"
)

# First Plot: Low Light (Dark & Artificial)
plot_dark <- ggplot(merged_data %>% filter(lighting_condition_group == "Low Light (Dark & Artificial)"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.7) +
  scale_fill_manual(values = c("Low Light (Dark & Artificial)" = "#54350a")) +  # Custom color
  labs(x = "Risk Index", y = "Density", title = "Low Light (Dark & Artificial)") +
  theme_minimal() +
  theme(legend.position = "none")

# Second Plot: Dawn and Artificial Light
plot_dawn <- ggplot(merged_data %>% filter(lighting_condition_group == "Dawn and Artificial Light"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Dawn and Artificial Light" = "#54350a")) +  # Custom color
  labs(x = "Risk Index", y = "Density", title = "Dawn and Artificial Light") +
  theme_minimal() +
  theme(legend.position = "none")

# Third Plot: Bright Daylight and Artificial Light
plot_daylight <- ggplot(merged_data %>% filter(lighting_condition_group == "Bright Daylight and Artificial Light"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Bright Daylight and Artificial Light" = "#54350a")) +  # Custom color
  labs(x = "Risk Index", y = "Density", title = "Bright Daylight and Artificial Light") +
  theme_minimal() +
  theme(legend.position = "none")

# Fourth Plot: Other Lighting Conditions
plot_other <- ggplot(merged_data %>% filter(lighting_condition_group == "Other Lighting Conditions"), aes(x = risk_index, fill = lighting_condition_group)) +
  geom_density(alpha = 0.6) +
  scale_fill_manual(values = c("Other Lighting Conditions" = "#54350a")) +  # Custom color
  labs(x = "Risk Index", y = "Density", title = "Other Lighting Conditions") +
  theme_minimal() +
  theme(legend.position = "none")

# Display the four plots side by side
plot_dark
plot_dawn
plot_daylight
plot_other
```

@fig-risk-index-lighting-conditions display the distribution of the Risk Index across various lighting conditions. The data is broken down into four main lighting categories: Low Light (Dark & Artificial), Dawn and Artificial Light, Bright Daylight and Artificial Light, and Other Lighting Conditions.

@fig-risk-index-lighting-conditions-1 highlights the risk distribution during low-light conditions, including both "Dark" and "Dark, Artificial" categories. The density curve shows that motorbike risks are relatively higher in these conditions, particularly in the moderate-to-high range of the Risk Index, indicating that poorer visibility and the presence of artificial lighting increase the likelihood of incidents.

@fig-risk-index-lighting-conditions-2 captures the risk profile during dawn and artificial lighting. The distribution shifts slightly, showing a lower concentration of higher risk indices compared to the low-light category. While there is still some risk during dawn, the overall risk index appears to be less extreme compared to the "Low Light" category, possibly due to the improving visibility as the sun rises.

@fig-risk-index-lighting-conditions-3 shows a much lower density of high-risk events. This suggests that the highest risk does not generally occur during daylight hours, as visibility is much better, and traffic conditions tend to be more predictable.

@fig-risk-index-lighting-conditions-4 captures a small number of cases that don’t fit into the above categories. The risk distribution in this category is less pronounced, likely because it includes a variety of edge cases, such as unusual lighting conditions or missing data.

### Neighborhood Risk Distribution

```{r}
#| label: fig-his_collosion_risk
#| fig-cap: "Neighborhoods grouped by average collision probabilities into High, Medium, and Low risk categories."
#| echo: false
#| warning: false
#| message: false
#| fig-width: 7
#| fig-height: 3.5
#|

# Load necessary libraries
library(tidyverse)
library(hrbrthemes)

# Load the dataset
data <- read.csv(here("data", "02-analysis_data", "collision_risk_scores.csv"))

# Take the average for duplicate neighborhoods
data <- data %>%
  group_by(hood_158) %>%
  summarise(collision_probability = mean(collision_probability, na.rm = TRUE)) %>%
  ungroup()

# Create groups: High, Medium, Low
data <- data %>%
  mutate(
    risk_group = case_when(
      collision_probability >= 0.8 ~ "High",
      collision_probability >= 0.6 ~ "Medium",
      TRUE ~ "Low"
    )
  )
